I would like to thank my advisor, Prof. Michael Reiter, for his support and
guidance throughout my Ph.D. study. I am sincerely grateful for his enthusiasm,
patience, and mentorship. I would also like to thank Prof. Cristian Cadar,
Prof. Somesh Jha Prof. Dinesh Manocha, and Prof. Fabian Monrose for graciously
serving on my dissertation committee. I am grateful for the time they devoted
to reading both my proposal and this dissertation; and for their many
questions, insightful comments, and helpful suggestions. I also
thank my research collaborators and co-authors, Darrell Bethea, Andrew Chi,
Marie Nesfield, and Prof. Cynthia Sturton as well as all other members of the
UNC security group.

I thank Dr. Ben Livshits for his mentorship at Microsoft Research and for the
opportunity to work on projects built upon symbolic reasoning. I also thank
Prof. Brian Dean, Prof. Robert Geist, and Prof. Jim Martin at Clemson
University for introducing me to computer science research and for encouraging
me to pursue graduate school.                                                      

I thank all my friends in the UNC Computer Science Department, particularly my
fellow bandidos: Darrell Bethea, David Millman, Brittany Fasy, and Catie Welsh.
I thank Jim Kuras and Lile Stephens for their friendship and artistic
inspiration.  I also sincerely thank everyone in the improv comedy community of
Chapel Hill.

Last but not least, I would like to thank my amazing family. I thank my
parents, Robert and Janeen Cochran for their constant encouragement throughout
my academic journey and their endless love and support. I also thank each of my
siblings: William, John Clark, Brooks, Harry, and Rose. I love each of you very
much and could not have finished this dissertation without you.

